A Review of Mortality Models

### 2020-09-14 mortality-models/draft.tex
WIP 14Sep2020

John Graunt, an English haberdasher, was perhaps the world's first demographer
and epidemiologist. In spite of the rudimentary mathematical tools available
in the seventeenth century, he was able to publish the first life tables. They
were based on the `Bills of Mortality' published in the parishes of London
every Thursday. The data was not best suited for constructing life tables but
that was the only one available in Graunt's time. Graunt admitted the 
inadequacy of his approach. Edmund Halley, the man after whom a comet is 
named, could get demographic data of a much better quality from the city of
Wroclaw in Poland. It had the age and sex of the dead individuals. Halley
used this data to build life tables of a greater accuracy than Graunt's. He
also demonstrated how to use the life tables to compute annuities. Graunt's
and Halley's work was entirely empirical, keeping with the times in which
it was done. Using their life tables was also quite laborious. With the 
invention of Calculus in the late seventeenth century and the development of 
the rudimentary ideas of a mathematical function, it became possible to build
theoretical models of human mortality.

@@

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalizations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10.

@@

Splines were used to draw a smooth curve passing through a set of points. They
were initially used by draftsmen who represented the points by heavy leads and
the splines by flexible wooden strips. This idea was borrowed into numerical
analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
of the twentieth century. A set of polynomials were used to approximate a
non-linear function where each polynomial fitted a portion of the function
being approximated. It was possible to show that under certian conditions
\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
piecewise polynomial function. The polynomials were required to blend into each
other that the boundaries in a smooth way. That is, not only were their values
required to match but also their derivatives.

Splines can be implemented using the Cox-de Boor recurrence relation
\cite{de1978practical}
\begin{equation}\label{s13e1}
B_{i,0}(x) = \begin{cases}
1 & \text{ if } x \in [t_i, t_{i+1}] \\
0 & \text{ otherwise}
\end{cases}
\end{equation}
and
\begin{equation}\label{s13e2}
B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
\end{equation}

>>>>

@@

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalizations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10. 
The models considered so far have looked at mortality irrespective of the
cause of death. However, in recent years, death rate due to certain causes
is falling while those due to others is rising. One can, therefore, forecast
death rates by cause and aggregate the results. This is the approach taken by
the cause-of-death models in section \ref{s11}. Several mortality models can
be written in the form $f(\mu) = L(\theta)$ where $L$ is a linear function of
the parameters $\theta$ of the model and $f$ is some function of the force
of mortality that permits us to have the right hand side as $L(\theta)$. The
statistically rigorous method to fit such models to mortality data is called
the generalized linear models. Section \ref{s12} discusses them. Not all
models can be brought in this format. Important examples being Lee-Carter and
Renshaw-Haberman. One has to take recourse to spline regression in these
cases. They are the topic of section \ref{s13}.

@@

Splines were used to draw a smooth curve passing through a set of points. They
were initially used by draftsmen who represented the points by heavy leads and
the splines by flexible wooden strips. This idea was borrowed into numerical
analysis by I. J. Shoenberg \cite{schoenberg1969cardinal} in the late sixties
of the twentieth century. A set of polynomials was used to approximate a
non-linear function where each polynomial fitted a portion of the function
being approximated. It was possible to show that under certian conditions
\cite{schoenberg1969cardinal} a non-linear function could be expressed as a 
piecewise polynomial function. The polynomials were required to blend into each
other that the boundaries in a smooth way. That is, not only were their values
required to match but also their derivatives.

Splines can be implemented using the Cox-de Boor recurrence relation
\cite{de1978practical}
\begin{equation}\label{s13e1}
B_{i,0}(x) = \begin{cases}
1 & \text{ if } x \in [t_i, t_{i+1}] \\
0 & \text{ otherwise}
\end{cases}
\end{equation}
and
\begin{equation}\label{s13e2}
B_{i,j}(x) = \frac{x - t_i}{t_{i+j} - t_i}B_{i,j-1}(x) + 
\frac{t_{i+j+1}-x}{t_{i+j+1}-t_{i+1}}B_{i+1,j-1}(x).
\end{equation}
These splines are then joined at the knot points $\{t_i: i = 1, \ldots, n\}$.
Once the knots are given it is easy to find the splines. However, there is
little guidance on the number of knots. Too few of them lead to under-fitting
and too many of them lead to over-fitting. O'Sullivan \cite{o1986statistical}
suggested that one should use a relatively large number of knots and prevent
over-fitting by levying a penalty on the second derivatives of the splines.
Eilers and Marx \cite{eilers1996flexible} simplify O'Sullivan's suggestion
and apply it to regression on B-splines. They applied the penalty to the
difference between the coefficients of the adjacent splines. The optimal amount
of smoothing is determined by cross-validation and the Akaike Information 
Criterion (AIC). Eilers and Marx call splines with penalties as P-splines.
They demonstrate that P-splines are an extension of the generalized linear
models. P-spline transforms conserve statistical moments of the data and have
polynomial curve fits as limits. Currie and others \cite{currie2004smoothing}
show how to use P-splines to fit mortality forecasting models. If $x_a = (
x_{a,1},\ldots,x_{a,n_a})^T$ and $x_y=(x_{y,1},\ldots,x_{y,n_y})^T$ are the
vectors with ages and years, $\mathcal{B}_a = \{B_{a,1}, \ldots, B_{a, c_a}\}$
and $\mathcal{B}_y = \{B_{y,1},\ldots,B_{y, c_y}\}$ are the cubic B-slines of
dimensions $c_a$ and $c_y$ then
\begin{equation}\label{c14e1}
X = B_y \otimes B_a
\end{equation}
is the regression matrix for the P-spline model
\begin{equation}\label{c14e2}
M(x,t) = \sum_{i,j}\theta_{i,j}B_{i,j}^{ay}(x, t).
\end{equation}
Here $B_a =(B_{a,j}(x_{a,i}))$ is an $n_a \times c_a$ matrix and $B_y = 
(B_{y,j}(x_{y,i}))$ is an $n_y \times c_y$ matrix. The penalty matrix is given
as
\begin{equation}\label{c14e3}
P = \lambda_aI_{c_y} \otimes D_a^TD_a + \lambda_y D_y^TD_y\otimes I_{c_a},
\end{equation}
where $D_a$ is the difference matrix of the rows of M, $D_y$ is the difference
matrix on the columns of M, and the two $\lambda$s are the smoothing parameters.
The regression problem is then minimizing the likelihood function
\begin{equation}\label{c14e4}
\ell_p = \ell - \frac{1}{2} m^TPm,
\end{equation}
where $\ell$ is the default log likelihood function and $m$ is the vector of
the elements of $M$.

The greatest benefit of splines over polynomial regression is that the former is
a more stable numerical method, especially when used for forecasting. The
reason for this stability is the local non-zero support for splines. 
Polynomials, on the other hand, have a global support. The P-spline regression
not only fits the mortality data smoothly, it also produces confidence intervals
to its forecasts enabling us to compute bounds for financial parameters 
depending on mortality.
